Light and life form interact to shape stomatal ratio among British angiosperms$^1$

### 2017-07-09 ms/ms.tex
put directory on github

% Notes from call with Rob
% van Tienderen - sensitivity approach better for evolution
% send list of species to rob

% TASKS
% what's going on with monocots?
% in methods - mention stuff about Salisbury
% leaf size

@@

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

<<Preliminaries, eval=TRUE, echo=FALSE, results=hide>>=

  # REMOVE BEFORE PUBLICATION
  system('cp ~/Google\\ Drive/CommonResources/refs.bib ~/Google\\ Drive/StomatalRatio/BEF/ms/refs.bib')

  # Directories
  pathMain <- "~/Google Drive/StomatalRatio/BEF"
  pathRawData <- paste0(pathMain, "/RawData")
  pathProcData <- paste0(pathMain, "/ProcData")
  pathR <- paste0(pathMain, "/R")
  pathMS <- paste0(pathMain, "/ms")
  pathFigures <- paste0(pathMS, "/Figures")

  # Libraries
  library(vioplot)
  library(ape)
  library(rncl)
  library(phytools)
  library(rstanarm)
  library(phylolm)
  library(Rphylopars)
  library(lavaan)
  library(stringr)
  library(magrittr)
  library(taxize)
  library(dplyr)

  # Source functions
  source(paste0(pathR, "/functions.R"))

@

<<Process data, eval=TRUE, echo=FALSE, results=hide>>=

  # Stomata
  ## BEF
  stomata <- read.csv(paste0(pathRawData, "/stomata.csv"), stringsAsFactors = F, 
                      strip.white = T)

  ## Salisbury 1927
  salisbury <- read.csv(paste0(pathRawData, "/salisbury_1927.csv"), 
                        stringsAsFactors = FALSE, strip.white = TRUE)

  # This is done as a loop to prevent timing out
  # tnr <- tnrs(query = salisbury$species[1], source = "iPlant_TNRS")
  # for (i in 2:nrow(salisbury)) {
  #  tnr %<>% rbind(tnrs(query = salisbury$species[i], source = "iPlant_TNRS"))
  # }
  # write.csv(temp, file = paste0(pathRawData, "/taxized_salisbury_1927.csv"), row.names = F)
  tnr <- read.csv(paste0(pathRawData, "/taxized_salisbury_1927.csv"), stringsAsFactors = FALSE)

  # When TNRS does not give an accepted name, I am using these based on searching The Plant List
  # http://www.theplantlist.org/
  # Plant List corrections

  old_name <- c("Betula alba",            "Betula pubescens",         "Fagus sylvatica",
                "Ulmus montana",          "Corylus avellana",         "Crataegus oxyacanthoides",
                "Rubus corylifolius",     "Cephalanthera pallens",    "Conopodium denudatum",
                "Helleborus foetidus",    "Primula acaulis",          "Byronia dioica",
                "Geranium phaeum",        "Geranium striatum",        "Hieracium sylvaticum",
                "Hypericum pulchrum",     "Lathyrus macrorrhizus",    "Luzula forsteri",
                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
                "Tilia parvifolia",       "Tilia vulgaris",           "Vicia sylvatica",
                "Iris foetidissima",      "Arrhenatherum avenaceum",  "Hordeum sylvaticum")
  new_name <- c("Betula pubescens",       "Betula pubescens",         "Fagus sylvatica",
                "Ulmus glabra",           "Corylus avellana",         "Crataegus laevigata",
                "Rubus plicatus",         "Cephalanthera longifolia", "Conopodium majus",
                "Helleborus foetidus",    "Primula vulgaris",         "Byronia dioica",
                "Geranium phaeum",        "Geranium versicolor",      "Hieracium murorum",
                "Hypericum pulchrum",     "Lathyrus linifolius",      "Luzula forsteri",
                "Luzula pilosa",          "Melittis melissophyllum",  "Sison amomum",
                "Tilia cordata",          "Tilia x europaea",         "Vicia sylvatica",
                "Xyridion foetidissimum", "Arrhenatherum elatius",    "Hordelymus europaeus")

  tnr$acceptedname[match(old_name, tnr$submittedname)] <- new_name
  tnr %<>% mutate(species = submittedname)

  salisbury %<>% inner_join(tnr, by = "species")
  rm(tnr)

  # Find which Salisbury data are already in BEF data
  already_in <- which(salisbury$submittedname %in% stomata$species |
                        salisbury$acceptedname %in% stomata$species)
  x1 <- salisbury$submittedname %in% stomata$species
  salisbury$species[x1] <- salisbury$submittedname[x1]
  x2 <- (salisbury$acceptedname %in% stomata$species) & !x1
  salisbury$species[x2] <- salisbury$acceptedname[x2]

  tmp <- inner_join(stomata, salisbury, by = "species")

  # Overlapping data from BEF and Salisbury are clearly taken from Salisbury, but with some errors. Hence, when they do not match, I have use the Salisbury data I entered myself.
  with(tmp, plot(ab_density.x, ab_density.y))
  with(tmp, plot(ad_density.x, ad_density.y))

  x <- tmp %>%
    use_series(ab_density.x) %>%
    equals(tmp %>% use_series(ab_density.y)) %>%
    not() %>%
    which()

  tmp$ab_density.x[x] <- tmp$ab_density.y[x]

  x <- tmp %>%
    use_series(ad_density.x) %>%
    equals(tmp %>% use_series(ad_density.y)) %>%
    not() %>%
    which()

  tmp$ad_density.x[x] <- tmp$ad_density.y[x]

  tmp %<>% 
    select(species, acceptedname, ab_density.x, ad_density.x) %>%
    set_colnames(c("species", "acceptedname", "ab_density", "ad_density"))

  stomata %<>% 
    left_join(tmp, by = c("species", "ab_density", "ad_density"))

  # Add data from Salisbury not in BEF
  stomata %<>% 
    rbind(salisbury[!(x1 | x2), colnames(.)])

  # Remove duplicates
  stomata <- stomata[!duplicated(stomata$species), ]

  # Calculate stomatal ratio
  get_sr_propAd <- function(X) {
    as.numeric(X["ad_density"]) / (as.numeric(X["ab_density"]) + as.numeric(X["ad_density"]))
  }
  get_sr_even <- function(X) {
    min(as.numeric(X[c("ab_density", "ad_density")])) / 
      max(as.numeric(X[c("ab_density", "ad_density")]))
  }
  stomata %<>% mutate(sr_propAd = apply(., 1, get_sr_propAd),
                      sr_even = apply(., 1, get_sr_even))

  stomata <- stomata[which(!is.na(stomata$sr_propAd)), ]

  # Photosynthetic pathway
  photo <- read.csv(paste0(pathRawData, "/photo.csv"), stringsAsFactors = F, 
                    strip.white = T)

  # All missing taxa are C3 
  # Not C4 based on Sage et al 2011 (JXB 62:9 3155-3169)
  # Not CAM based on photosynthetic pathway of congeneric species in the data
  missing_photo <- 
    c("Alchemilla vulgaris",  "Asarum europaeum", "Byronia dioica",
      "Hieracium sylvaticum", "Luzula maxima",    "Prunus cerasus",
      "Pteridium aquilinum",  "Pyrus communis",   "Pyrus torminalis",
      "Quercus sessiliflora", "Rhamnus frangula", "Ribes grossularia",
      "Rubus corylifolius",   "Tilia vulgaris")
  photo %<>% rbind(data.frame(species = missing_photo, photo = "c3"))
  all(stomata$species %in% photo$species |
        stomata$acceptedname %in% photo$species) # should be true

  # PLANTATT (Hill et al 2004) for lifeform and Ellenberg light indicator values
  plantatt <- read.csv(paste0(pathRawData, "/plantatt.csv"), stringsAsFactors = F, 
                       strip.white = T)
  # stomata$species[!(stomata$species %in% plantatt$Taxon.name |
  #                    stomata$acceptedname %in% plantatt$Taxon.name)] # Missing data for several

  # Combine bulbous and nonbulbous geophytes
  plantatt$LF1[plantatt$LF1 == "Gb"] <- "Gn"

  # Combine hydrophytes and annual hydrophytes
  plantatt$LF1[plantatt$LF1 == "Hz"] <- "Hy"

  # Combine phanerophytes and nanophanerophytes
  plantatt$LF1[plantatt$LF1 == "Pn"] <- "Ph"

  ##### MESSING AROUND --------------------------------------------
  # Leaf size
  leaf_size <- read.csv(paste0(pathRawData, "/leaf_size.csv"), stringsAsFactors = F)
  leaf_size$leaf_size %<>% ordered(levels = c("<0.1", "0.1-1", "1-10", "10-100", "100-1000",
  																						">1000"))
  leaf_size <- read.csv(paste0(pathRawData, "/leda_leaf_area.csv"), stringsAsFactors = F,
  											strip.white = TRUE)
  leaf_size %<>% rename(species = SBS.name, leaf_size = single.value..mm.2.)
  leaf_size <- data.frame(leaf_size = tapply(leaf_size$leaf_size, leaf_size$species, mean))
  leaf_size$species <- rownames(leaf_size)

  seed_number <- read.csv(paste0(pathRawData, "/leda_seed_number.csv"), stringsAsFactors = F,
  											strip.white = TRUE)
  seed_number %<>% rename(species = SBS.name, seed_number = single.value)
  seed_number <- data.frame(seed_number = tapply(seed_number$seed_number, seed_number$species, mean))
  seed_number$species <- rownames(seed_number)
  which(!(stomata$species %in% seed_number$species |
  			stomata$acceptedname %in% seed_number$species))

  #####-----------------------------

  # Many missing leaf size
  which(!(stomata$species %in% leaf_size$species |
  			stomata$acceptedname %in% leaf_size$species))

  # Combine datasets
  xp <- match(stomata$species, photo$species)
  xpa <- match(stomata$acceptedname, photo$species)
  xp[is.na(xp)] <- xpa[is.na(xp)]

  xa <- match(stomata$species, plantatt$Taxon.name)
  xaa <- match(stomata$acceptedname, plantatt$Taxon.name)
  xa[is.na(xa)] <- xaa[is.na(xa)]

  xl <- match(stomata$species, leaf_size$species)
  xla <- match(stomata$acceptedname, leaf_size$species)
  xl[is.na(xl)] <- xla[is.na(xl)]

  xs <- match(stomata$species, seed_number$species)
  xsa <- match(stomata$acceptedname, seed_number$species)
  xs[is.na(xs)] <- xsa[is.na(xs)]

  stomata$photo <- photo$photo[xp]
  stomata$lifeform <- plantatt$LF1[xa]
  stomata$ellenberg_light <- plantatt$L[xa]
  stomata$height <- plantatt$Hght[xa]
  stomata$wh <- plantatt$W[xa]
  stomata$leaf_size <- leaf_size$leaf_size[xl]
  stomata$seed_number <- seed_number$seed_number[xs]

  # Remove missing values
  stomata <- stomata[complete.cases(stomata[, c("sr_propAd", "lifeform", "ellenberg_light")]), ]

  rm("already_in", "missing_photo", "new_name", "old_name", "photo", 
     "plantatt", "salisbury", "tmp", "x", "x1", "x2", "xp", "xpa", "xa", "xaa")

  # Export  
  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))

@

<<Leaf size, eval=TRUE, echo=FALSE, results=hide>>=

  # boxplot(sr_even ~ leaf_size, data = stomata)
  # boxplot(ellenberg_light ~ leaf_size, data = stomata)

  plot(sr_even ~ leaf_size, data = stomata, log = "x")
  boxplot(log10(leaf_size) ~ ellenberg_light, data = stomata)
  tmp <- read.csv(paste0(pathRawData, "/try_leaf_area.csv"), stringsAsFactors = FALSE,
  								strip.white = TRUE)

  stomata$species %in% tmp$AccSpeciesName | stomata$acceptedname %in% tmp$AccSpeciesName
 sort(table(tmp[which(tmp$AccSpeciesName %in% stomata$species), c("Dataset")]))
@

<<Phylogeny, eval=TRUE, echo=FALSE, results=hide>>=

  # Import phylogeny of British Flora from Lim et al 2014
  tmp <- readLines(paste0(pathRawData, "/S15105.nex"))
  l1 <- grep("BEGIN CHARACTERS", tmp) - 1
  l2 <- grep("BEGIN TREES", tmp) - 1
  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathRawData, "/Lim_etal_2014.nex"))
  phy <- read_nexus_phylo(paste0(pathRawData, "/Lim_etal_2014.nex"))

  # Check for species missing from phylogeny
  stomata$species %<>% str_replace_all(" ", "_")
  stomata$acceptedname %<>% str_replace_all(" ", "_")
  notInPhy <- which(!(stomata$species %in% phy$tip.label |
                        stomata$acceptedname %in% phy$tip.label))
  stomata[notInPhy, ]; length(notInPhy)

  # Modify tip labels

  # Using Cerastium brachypetalum as proxy for C. cerastoides based on similar divergence in Scheen et al. 2004 (Am J Bot, 91(6): 943-952)
  tmp <- gsub("Cerastium_brachypetalum", "Cerastium_cerastoides", tmp)

  # Using Coincya monensis as proxy for C. wrightii 
  tmp <- gsub("Coincya_monensis", "Coincya_wrightii", tmp)

  # Using Geranium pusillum as proxy for G. versicolor based on similar divergence in Palazzesi et al. 2012 (Bio J of Linn Soc, 107(1): 67-85)
  # tmp <- gsub("Geranium_pusillum", "Geranium_versicolor", tmp)

  # Using Gnaphalium luteoalbum as proxy for G. norvegicum 
  tmp <- gsub("Gnaphalium_luteoalbum", "Gnaphalium_norvegicum", tmp)

  # Lithospermum purpurocaeruleum is mispelled
  tmp <- gsub("Lithospermum_purpureocaeruleum", "Lithospermum_purpurocaeruleum", tmp)

  # Using Lythrum portula as proxy for L. hyssopifolia 
  tmp <- gsub("Lythrum_portula", "Lythrum_hyssopifolia", tmp)

  # Using Peucedanum ostruthium as proxy for P. officinale 
  tmp <- gsub("Peucedanum_ostruthium", "Peucedanum_officinale", tmp)

  # Changing Populus_nigra_sens.lat. to Populus nigra officinale 
  tmp <- gsub("Populus_nigra_sens.lat.", "Populus_nigra", tmp)

  # Changing Rosa_canina_agg. to Rosa_canina officinale 
  tmp <- gsub("Rosa_canina_agg.", "Rosa_canina", tmp)

  # Using Stachys alpina as proxy for S. germanica based on similar divergence in Salmaki et al. 2013 (Mol Phylo Evol, 69(3): 535-551)
  tmp <- gsub("Stachys_alpina", "Stachys_germanica", tmp)

  # Using Tripleurospermum_maritimum as proxy for T. inodorum 
  tmp <- gsub("Tripleurospermum_maritimum", "Tripleurospermum_inodorum", tmp)

  # Using Viola_arvensis as proxy for V. kitaibeliana 
  tmp <- gsub("Viola_arvensis", "Viola_kitaibeliana", tmp)

  # Write first modified phylogeny
  writeLines(tmp[c(1:l1, l2:length(tmp))], con = paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))
  phy <- read_nexus_phylo(paste0(pathProcData, "/Lim_etal_2014_mod1.nex"))

  # Bind additional tips
  tips <- phy$tip.label
  nodes <- sapply(tips, function(x,y) which(y == x), y = phy$tip.label)
  edge.lengths <- setNames(phy$edge.length[sapply(nodes, function(x, y) {
    which(y == x) }, y = phy$edge[,2])], names(nodes))

  # Make Agrostemma githago sister to all (Silene sp., Lychnis sp.) [based on Fior et al. 2006, Am J Bot 104(2):399-411]
  n <- getMRCA(phy, phy$tip.label[c(grep("Silene", phy$tip.label), grep("Lychnis", phy$tip.label))])

  # Branch length is average of distance from (Silene sp., Lychnis sp.) MRCA to (Silene sp., Lychnis sp.) sp.
  el1 <- sapply(nodes[c(grep("Silene_", phy$tip.label), grep("Lychnis_", phy$tip.label))], 
                nodeheight, tree = phy)
  el2 <- nodeheight(tree = phy, node = n)

  phy <- bind.tip(phy, "Agrostemma_githago", edge.length = mean(el1 - el2), where = n)

  # Make Heracleum mantegazzianum as sister to H. sphondylium
  phy <- bind.tip(phy, "Heracleum_mantegazzianum", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Heracleum_sphondylium")],
                  where = which(phy$tip.label == "Heracleum_sphondylium"))

  # Make Portulaca oleracea sister to all Claytonia (based on Angiosperm Phylogeny Website v 13 http://www.mobot.org/mobot/research/apweb/welcome.html]
  # n <- getMRCA(phy, phy$tip.label[grep("Claytonia", phy$tip.label)])

  # Branch length is average of distance from Claytonia MRCA to Claytonia sp.
  # el1 <- sapply(nodes[grep("Claytonia_", names(nodes))], nodeheight, tree = phy)
  # el2 <- nodeheight(tree = phy, node = n)

  # phy <- bind.tip(phy, "Portulaca_oleracea", edge.length = mean(el1 - el2), where = n)

  # Make Spartina alterniflora, Spartina maritima, and Spartina x townsendii as sister to S. anglica
  phy <- bind.tip(phy, "Spartina_alterniflora", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
                  where = which(phy$tip.label == "Spartina_anglica"))

  phy <- bind.tip(phy, "Spartina_maritima", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
                  where = which(phy$tip.label == "Spartina_anglica"))

  # phy <- bind.tip(phy, "Spartina_x_townsendii", 
  #                 edge.length = phy$edge.length[which(phy$tip.label == "Spartina_anglica")],
  #                 where = which(phy$tip.label == "Spartina_anglica"))

  # Make Tephroseris integrifolia sister to all Senecio sp.
  n <- getMRCA(phy, phy$tip.label[grep("Senecio", phy$tip.label)])

  # Branch length is average of distance from Senecio MRCA to Senecio sp.
  el1 <- sapply(nodes[grep("Senecio_", names(nodes))], nodeheight, tree = phy)
  el2 <- nodeheight(tree = phy, node = n)

  phy <- bind.tip(phy, "Tephroseris_integrifolia", edge.length = mean(el1 - el2), where = n)

  # Make Viola hirta as sister to V. reichenbachiana (both in Viola sect. Viola)
  phy <- bind.tip(phy, "Viola_hirta", 
                  edge.length = phy$edge.length[which(phy$tip.label == "Viola_reichenbachiana")],
                  where = which(phy$tip.label == "Viola_reichenbachiana"))

  # Write second modified phylogeny
  write.nexus(phy, file = paste0(pathProcData, "/Lim_etal_2014_mod2.nex"))

  # Check for species missing from modified phylogeny
  notInPhy <- which(!(stomata$species %in% phy$tip.label |
                        stomata$acceptedname %in% phy$tip.label))
  stomata[notInPhy, ]; length(notInPhy)

  rm("edge.lengths", "el1", "el2", "l1", "l2", "n", "nodes", "notInPhy", "tips", "tmp")

  # Run PATHd8 (Britton et al. 2002)
  cat(sprintf("Sequence length = 4692;

              %s

              mrca: Azolla_filiculoides, Solanum_nigrum, maxage=454;
              mrca: Osmunda_regalis, Asplenium_onopteris, minage=354;
              mrca: Pinus_sylvestris, Solanum_nigrum, minage=310;
              mrca: Nymphaea_alba, Solanum_nigrum, minage=140;
              mrca: Nymphaea_alba, Solanum_nigrum, maxage=217;
              mrca: Nymphaea_alba, Nuphar_lutea, minage=115;
              mrca: Chelidonium_majus, Solanum_nigrum, fixage=121;
              mrca: Quercus_robur, Bryonia_dioica, minage=84;

              name of mrca: Azolla_filiculoides, Solanum_nigrum, name=tracheophytes;
              name of mrca: Osmunda_regalis, Asplenium_onopteris, name=monilophytes;
              name of mrca: Pinus_sylvestris, Solanum_nigrum, name=seedplants;
              name of mrca: Nymphaea_alba, Solanum_nigrum, name=angiosperms;
              name of mrca: Nymphaea_alba, Nuphar_lutea, name=nymphaceae;
              name of mrca: Chelidonium_majus, Solanum_nigrum, name=eudicots;
              name of mrca: Quercus_robur, Bryonia_dioica, name=curcibitales_fagales;

              ", write.tree(phy)), file = paste0(pathMain, "/PATHd8/pathd8_in"))

  # Run PATHd8
  system("cd ~/Google\\ Drive/StomatalRatio/BEF/PATHd8; ./PATHd8 pathd8_in pathd8_out")

  # Import ultrametric tree
  pathd8_out <- readLines(paste0(pathMain, "/PATHd8/pathd8_out"))
  phy <- read.tree(text = substr(pathd8_out[11], 14, nchar(pathd8_out[11])))

  rm("pathd8_out")

@

<<Violin plot of Raunkiaer life form versus stomatal ratio, eval=TRUE, echo=FALSE, results=hide>>=

  # chamaephyte = subshrub (woody)
  # geophyte = cryptophyte resting in dry ground
  # hemicryptophyte = perennial with overwintering buds at soil surface
  # hydrophyte = cryptophyte resting in marshy ground
  # phanerophyte = normally woody perennial
  # therophyte = annual

  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "hydrophyte", "phanerophyte", "therophyte")
  names(lf) = c("Ch", "Gn", "hc", "Hy", "Ph", "Th")
  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]

  pdf(paste0(pathFigures, "/FigureS_violin.pdf"), 8, 5.25)
  par(mai = c(1.75, 1.5, 0.25, 0.25), cex.lab = 1.5, las = 1, cex = 1.5, cex.axis = 0.5)

  vioplot(subset(stomata$sr_propAd, stomata$lifeform == names(lf[1]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[2]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[3]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[4]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[5]), na.rm = TRUE),
          subset(stomata$sr_propAd, stomata$lifeform == names(lf[6]), na.rm = TRUE),
          col = "grey", colMed = "black", names = rep("", 6))

  clip(grconvertX(0, "ndc", "user"), grconvertX(1, "ndc", "user"),
       grconvertY(0, "ndc", "user"), grconvertY(1, "ndc", "user"))
  # Sample sizes
  text(1:6 + 0.25, grconvertY(1.5, "inches", "user"), labels = lf, srt = 30, pos = 2)
  mtext(table(stomata$lifeform)[names(lf)], 3, at = 1:6, line = 0)
  title(ylab = "Stomatal Ratio",  line = 3.5)
  title(xlab = "Raunkiaer Life Form", line = 4.5)
  title(ylab = expression(SR[propAd] == SD[ad] / SD[total]), line = 2, cex.lab = 1) 
  dev.off()

  rm("lf")

@

<<Remove hydrophytes + c4 + cam + nonangiosperm, eval=TRUE, echo=FALSE, results=hide>>=

  nHydrophyte <- length(which(stomata$lifeform == "Hy"))
  nC4 <- length(which(stomata$photo == "c4"))
  nCAM <- length(which(stomata$photo == "cam"))

  x1 <- c("Ph", "Ch", "hc", "Gn", "Th")
  stomata <- stomata[which(stomata$lifeform %in% x1 & stomata$photo == "c3"), ]

  angioPhy <- extract.clade(phy, 
                            node = length(phy$tip.label) + which(phy$node.label == "angiosperms"))

  nNonangio <- length(which(!(stomata$species %in% angioPhy$tip.label | 
                                stomata$acceptedname %in% angioPhy$tip.label)))

  stomata <- stomata[stomata$species %in% angioPhy$tip.label |
                       stomata$acceptedname %in% angioPhy$tip.label, ]

  # Export new dataset and tree
  write.csv(stomata, paste0(pathProcData, "/stomata.csv"))
  write.nexus(angioPhy, file = paste0(pathProcData, "/angioPhy.nex"))

  rm("x1", "phy")

@

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

I excluded data from hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants.. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species. The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.

@@

<<Raunkiaer life form versus Ellenberg light indicator values, eval=TRUE, echo=FALSE, results=hide>>=

  # Make dichotomous and slightly adjust edge lengths to make exactly ultrametric
  modAngioPhy <- multi2di(angioPhy)
  modAngioPhy$edge.length[modAngioPhy$edge.length == 0] <- 0.02
  x <- diag(vcv.phylo(modAngioPhy))
  is.ultrametric(modAngioPhy)
  n <- length(modAngioPhy$tip.label)
  te <- sapply(1:n, function(x, y) which(y == x), y = modAngioPhy$edge[, 2]) # terminal edges
  modAngioPhy$edge.length[te] <- modAngioPhy$edge.length[te] + (max(x) - x)
  is.ultrametric(modAngioPhy)

  x <- which(stomata$acceptedname %in% modAngioPhy$tip.label &
               !(stomata$species %in% modAngioPhy$tip.label))
  stomata$species[x] <- stomata$acceptedname[x]

  x <- stomata$species[duplicated(stomata$species)]
  stomata[stomata$species %in% x, ]
  modAngioPhy <- drop.tip(modAngioPhy, 
                          modAngioPhy$tip.label[!modAngioPhy$tip.label %in% stomata$species])

  # phylolm shows that there is no phylogenetic signal in this regression (very high alpha)
  rownames(stomata) <- stomata$species # need to change row names for phylolm
  fitLFvEL_pl <- phylolm(ellenberg_light ~ -1 + lifeform, model = "OUrandomRoot", data = stomata, 
                         phy = modAngioPhy, upper.bound = 1e10)
  summary(fitLFvEL_pl)

  # therefore using standard ANOVA
  fitLFvEL <- lm(ellenberg_light ~ lifeform, data = stomata)
  aovLFvEL <- anova(fitLFvEL)

  # for plotting confidence intervals
  fitLFvEL <- lm(ellenberg_light ~ -1 + lifeform, data = stomata)

  lf <- c("chamaephyte", "geophyte", "hemicryptophyte", "phanerophyte", "therophyte")
  names(lf) = c("Ch", "Gn", "hc", "Ph", "Th")
  lf <- lf[names(sort(tapply(stomata$sr_propAd, stomata$lifeform, mean)))]

  pdf(paste0(pathFigures, "/Figure_lf-light.pdf"), 4, 7)
  par(mai = c(1.5, 1, 0, 0.25), cex.lab = 1, las = 1)
  plot(0, 0, type = "n", xlim = c(0, 9), ylim = c(0, 5), xlab = "", ylab = "", axes = F)

  for (i in 0:4) abline(h = i + seq(0.2, 0.6, 0.2), col = "grey")
  hs <- tapply(stomata$ellenberg_light, stomata$lifeform, hist, plot = F, breaks = 0:9)  
  for (i in 1:5) {
    rect(0:8, rep(i - 1, 9), 1:9, i - 1 + hs[[names(lf)[i]]]$density, col = "black", 
         border = "white", lwd = 2)
  }

  # label life forms
  s <- sprintf("%s (%s)", lf, table(stomata$lifeform)[names(lf)])
  text(0 + 0.5 * strwidth(s), 1:5, labels = s, pos = 1)

  # axes
  axis(1, at = 0.5:8.5, labels = 1:9, lwd = 0, line = -1.5)
  for (i in 0:4) axis(2, at = i + seq(0.2, 0.6, 0.2), labels = seq(0.2, 0.6, 0.2), lwd = 0, 
                      lwd.ticks = 1)

  # axis labels
  mtext(c("Shade\n", "Partial\nShade", "Sun\n"), 1, at = c(1.5, 4.5, 7.5), line = 1.5)
  title(xlab = "Ellenberg light\nindicator", cex.lab = 1.5, line = 5)
  title(ylab = "Proportion", cex.lab = 1.5)

  clip(grconvertX(0, from = "ndc", "user"), grconvertX(1, from = "ndc", "user"),
       grconvertY(0, from = "ndc", "user"), grconvertY(1, from = "ndc", "user"))
  rect(grconvertX(0, from = "npc", "user"), 0:4, 
       grconvertX(1, from = "npc", "user"), 0.75:4.75)

  # Means + 95% confidence intervals
  ci <- confint(fitLFvEL)
  ci <- ci[paste0("lifeform", names(lf)), ] # reorder
  mu <- coef(fitLFvEL)[paste0("lifeform", names(lf))]
  for (i in 1:5) {
    lines(ci[i, ], rep(i - 1 + 0.675, 2), lwd = 2)
    points(mu[i], i - 1 + 0.675, pch = 21, col = "black", bg = "white")
  }

  dev.off()

  rm("ci", "fitLFvEL_pl", "hs", "i", "mu", "n", "s", "te", "x")

@

@@

<<Raunkiaer life form and Ellenberg light indicator values versus SR_even, eval=TRUE, echo=FALSE, results=hide>>=

  # Phylogenetic regression model comparison
  fitSRa <- phylolm(sr_even ~ lifeform * ellenberg_light, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRb <- phylolm(sr_even ~ lifeform + ellenberg_light, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRc <- phylolm(sr_even ~ ellenberg_light, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRd <- phylolm(sr_even ~ lifeform, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  fitSRe <- phylolm(sr_even ~ 1, model = "OUrandomRoot",
                    data = stomata, phy = modAngioPhy, upper.bound = 10)
  aicSR <- sapply(list(fitSRa, fitSRb, fitSRc, fitSRd, fitSRe), AIC)

  # Parametric bootstrap CIs of full model for p-values
  # set.seed(36502)
  # fitSR_pl <- phylolm(sr_even ~ -1 + lifeform + ellenberg_light:lifeform, model = "OUrandomRoot",
  #                    data = stomata, phy = modAngioPhy, boot = 1e4, upper.bound = 10)
  # saveRDS(fitSR_pl, file = paste0(pathR, "/fitSR_pl.rds"))
  fitSR_pl <- readRDS(file = paste0(pathR, "/fitSR_pl.rds"))

  # Figure

  pdf(paste0(pathFigures, "/Figure_SRmultReg.pdf"), 4, 7)
  par(mfrow = c(5, 1), mar = rep(0, 4), cex.lab = 1, oma = c(5, 7, 1, 1))

  for (i in 5:1) {

    plot(0, 0, type = "n", xlim = c(1, 9), ylim = c(-0.05, 1.05), xlab = "", ylab = "", axes = F,
         frame.plot = F)
    if (i %in% seq(1, 5, 2)) axis(2, at = seq(0, 1, 0.5), lwd = 0, lwd.ticks = 1, las = 1)

    # polygon of confidence intervals
    x <- seq(min(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]),
             max(stomata$ellenberg_light[stomata$lifeform == names(lf)[i]]), length.out = 1e3)
    bs <- sapply(x, function(X) {
      fitSR_pl$bootstrap[, sprintf('lifeform%s', names(lf)[i])] + 
        fitSR_pl$bootstrap[, sprintf('lifeform%s:ellenberg_light', names(lf)[i])] * X
    } )
    ci <- apply(bs, 2, quantile, probs = c(0.025, 0.5, 0.975))
    polygon(c(x, rev(x)), c(ci[1, ], rev(ci[3, ])), col = "grey", border = NA)
    points(x, ci[2, ], lwd = 2, type = "l") # median
    points(x, ci[1, ], lwd = 2, lty = 2, type = "l") # lower 95%
    points(x, ci[3, ], lwd = 2, lty = 2, type = "l") # upper 95%

    # plot points
    with(subset(stomata, stomata$lifeform == names(lf)[i]), 
         points(jitter(ellenberg_light), sr_even, pch = 21, bg = rgb(0, 0, 0, 0.1), cex = 2))

>>>>

% TASKS
% what's going on with monocots?
% in methods - mention stuff about Salisbury

@@

\usepackage{Sweave}
\begin{document}
\input{ms-concordance}

@@

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 374 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

I excluded data from hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (3 species) and CAM (2 species) plants.. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 374 species. The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.

@@

@@
